Transfer learning for echelle spectroscopy

### 2021-11-15 paper/paper1/ms.tex
Writing about minibatches and sparsity

$\mathsf{V}_j(\bm{\lambda}_s)$ matrix  .

>>>>

Equation \ref{equation1} performs a type of tensor contraction, turning a $N_{\mathrm{lines}}\times N_{s}$ matrix into a $1\times N_{s}$ vector.  The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix, with a scalar prefactor for the cost of evaluating a single line profile at a single wavelength point.  Efficient GPU algorithms exist for tensor contractions such as this product along an axis of a matrix, allowing this computation to proceed quickly on modern machines\authorcomment1{add citation?}. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math, although the algorithms place restrictions on floating precision depending on the particular hardware and software available.  The chief bottleneck occurs when the storage of the $\bm{\bar{V}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM.  The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{V}}$ matrix, since the CPU/GPU also has to store the computation graph for autodiff.  The computation graph takes up more memory than the matrix itself, so it is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop.  A remedy is needed.

Minibatches offer a natural solution to the line-by-line computational
bottleneck\footnote{The Machine Learning and Artificial Intelligence (ML/AI) community employs minibatches to handle similar situations, in which the dataset volume exceeds the available GPU RAM.  Our configuration departs slightly from the typical ML/AI situation, since their models are usually ``small'' relative to their datasets.  Here our model is ``big'' relative to our dataset.}.
You can choose to assemble and evaluate only a portion of the matrix at a time.  The choice to evaluate only a portion of lines at a time would mean the model is inaccurately evaluated at those points.  Instead we choose to evaluate only a subset $N_{\mathrm{subset}}$ of the total pixels $N_s$, so that the model is exact at those points.  All lines update at each glimpse of a minibatch, but many lines get little information about how the loss scalar changes for their parameters.  The value of $N_{\mathrm{subset}}$ should be set just below the threshold at which the computation runs out of memory.  At present, this threshold is determined experimentally.  The indices of the wavelength points can either be set to be contiguous in blocks, or random with possibly large gaps between indices, or else with some other scheme.  We chose random sampling with replacement, meaning that some pixels get revisited more often than others.

Another solution to the computation problem could be to take advantage of the sparsity of the $\bm{\bar{V}}$ matrix: most of the entries are vanishingly close to zero.  Ideally we would only evaluate the model close to the line center position, some number of line cuts $N_{\mathrm{cut}}=\alpha \cdot \mathrm{FWHM}$ away.  These line cuts produce a speedup by a factor of $\frac{N_s}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra. Furthermore, efficient algorithms for assembling coalescing sparse matrices exist in \texttt{PyTorch}.  A few practical considerations make sparse matrices peskier to work with than their dense counterparts.  First, the $\alpha$ hyperparameter should be set large enough that the truncation effect is not seen for the deepest lines.  The $\alpha$ should be preselected before the training process.  If the FWHM initialization was poor enough, the question may arise how autodiff training will cope with the poor performance of a model, leading to the prospect of strange and difficult to diagnose artifacts near spurious line-cut truncation artifacts.  At the time of writing, we do not implement sparse matrices in \emph{blas\'e}, but we find them the most compelling route for transformative performance improvements.
